---
title: "Exploratory Data Analysis"
output:
  pdf_document:
    toc: false
    fig_crop: false
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagestyle{empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.pos = "H", out.extra = "", fig.align = "center")

library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
library(lubridate)
library(corrplot)
```

```{r load-data}
df <- read_csv("Buy_Now_Pay_Later_BNPL_CreditRisk_Dataset.csv")

df <- df %>%
  mutate(
    default_flag = factor(default_flag, levels = c(0, 1),
                          labels = c("No Default", "Default")),
    employment_type  = factor(employment_type),
    product_category = factor(product_category),
    location         = factor(location),
    bnpl_installments = factor(bnpl_installments),
    transaction_date = mdy(transaction_date)
  )
```
\thispagestyle{empty}
# Summary of Key EDA Findings

- Overall default rate is **39.1%**, implying a moderate data imbalance, which should be addressed by applying SMOTE.
- **Unemployed and Student users** have the highest default rates (~49%), nearly double that of Self-Employed users (~26%).
- Users with **lower credit scores** (300--550 range) have higher default probability.
- The default users have more **missed payments and repayment delays**.
- **Higher debt-to-income ratio** leads to higher default probability.
- **monthly_income** and **credit_score** exhibit a very high correlation (0.96), indicating multicollinearity; therefore, **monthly_income** is removed in the preprocessing stage.
- **customer_segment** and **risk_score** are removed in the preprocessing stage due to potential data leakage concerns.

# Dataset Overview

```{r dataset-overview}
overview <- tibble(
  Item  = c("Total Records", "Total Variables", "Defaulted Users",
            "Non-Defaulted Users", "Default Rate", "Missing Values"),
  Value = c(
    format(nrow(df), big.mark = ","),
    ncol(df),
    format(sum(df$default_flag == "Default"), big.mark = ","),
    format(sum(df$default_flag == "No Default"), big.mark = ","),
    paste0(round(mean(df$default_flag == "Default") * 100, 1), "%"),
    "None"
  )
)

kable(overview, caption = "Dataset Overview",
      booktabs = TRUE, align = c("l", "r")) %>%
  kable_styling(latex_options = c("hold_position"), full_width = FALSE)
```

**Interpretation:** The dataset contains 10,345 transactions across 17 variables. Overall, 39.1% of users defaulted on their repayments.

# Descriptive Statistics of Numerical Variables

```{r descriptive-stats}
num_vars <- df %>%
  select(age, monthly_income, credit_score, purchase_amount,
         repayment_delay_days, missed_payments,
         debt_to_income_ratio, risk_score, app_usage_frequency)

desc_stats <- num_vars %>%
  summarise(across(everything(), list(
    Mean   = ~round(mean(.), 2),
    Median = ~round(median(.), 2),
    SD     = ~round(sd(.), 2),
    Min    = ~round(min(.), 2),
    Max    = ~round(max(.), 2)
  ))) %>%
  pivot_longer(everything(),
               names_to  = c("Variable", "Statistic"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Statistic, values_from = value)

kable(desc_stats,
      caption = "Descriptive Statistics of Numerical Variables",
      booktabs = TRUE, align = c("l", rep("r", 5))) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

**Interpretation:**

- **Monthly Income** ranges widely, with extreme high values identified as outliers that warrant further attention in preprocessing.
- **Credit Score** ranges from 300 to 850, with a median of 403, suggesting many users have low creditworthiness.
- **Repayment Delay Days** averages 8.7 days, with some users delaying up to 33 days. **Missed Payments** ranges up to 7, and **Debt-to-Income Ratio** reaches 0.72. The extreme values in these three variables are removed in the preprocessing stage.

# Target Variable Distribution

```{r target-distribution, fig.width=4, fig.height=3}
df %>%
  count(default_flag) %>%
  mutate(pct = n / sum(n),
         label = paste0(comma(n), "\n(", percent(pct, accuracy = 0.1), ")")) %>%
  ggplot(aes(x = default_flag, y = n, fill = default_flag)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("No Default" = "#2196F3", "Default" = "#F44336")) +
  scale_y_continuous(labels = comma, limits = c(0, 7500)) +
  labs(title = "Distribution of Default Status",
       x = "Default Status", y = "Number of Users") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** The dataset has a moderate data imbalance problem. 61% of users did not default, while 39% did. SMOTE will be applied in data preprocessing stage to deal with this problem.

# Default Rate by Employment Type

```{r employment-default, fig.width=4.5, fig.height=3}
df %>%
  group_by(employment_type) %>%
  summarise(
    total   = n(),
    default = sum(default_flag == "Default"),
    rate    = default / total
  ) %>%
  ggplot(aes(x = reorder(employment_type, -rate), y = rate, fill = employment_type)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_y_continuous(labels = percent, limits = c(0, 0.6)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Default Rate by Employment Type",
       x = "Employment Type", y = "Default Rate") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Unemployed users (48.7%) and Students (48.6%) have nearly double the default rate of Self-Employed users (26.2%) and significantly higher than Salaried workers (32.7%). This is because users with unstable or no income face greater difficulty in paying installment punctually.

# Credit Score Distribution by Default Status

```{r credit-score-pyramid, fig.width=5, fig.height=3.5}
pyramid_data <- df %>%
  mutate(score_band = cut(credit_score,
                          breaks = seq(300, 850, by = 50),
                          include.lowest = TRUE,
                          right = FALSE)) %>%
  count(score_band, default_flag) %>%
  mutate(n_plot = ifelse(default_flag == "No Default", -n, n))

x_max <- max(abs(pyramid_data$n_plot), na.rm = TRUE) * 1.05

ggplot(pyramid_data, aes(x = n_plot, y = score_band, fill = default_flag)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.5) +
  scale_x_continuous(
    labels = function(x) comma(abs(x)),
    breaks = seq(-1200, 1200, by = 400),
    limits = c(-x_max, x_max)
  ) +
  scale_fill_manual(values = c("No Default" = "#2196F3", "Default" = "#F44336")) +
  labs(title = "Credit Score Distribution by Default Status",
       x = "Number of Users", y = "Credit Score Band") +
  annotate("text", x = -x_max * 0.55, y = 11.7,
           label = "No Default", colour = "#2196F3", fontface = "bold", size = 3.5) +
  annotate("text", x =  x_max * 0.55, y = 11.7,
           label = "Default",    colour = "#F44336", fontface = "bold", size = 3.5) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 11) +
  theme(plot.title    = element_text(face = "bold", hjust = 0.5),
        plot.margin   = margin(t = 20, r = 10, b = 10, l = 10),
        legend.position = "none")
```

**Interpretation:** Users who defaulted tend to have a low credit score, ranging from 300 to 550, while non-defaulters show a wider and slightly higher distribution. This means that lower credit scores are associated with higher default risk.

# Missed Payments & Repayment Delay vs. Default

```{r missed-payments, fig.width=4, fig.height=3}
ggplot(df, aes(x = default_flag, y = missed_payments, fill = default_flag)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("No Default" = "#2196F3", "Default" = "#F44336")) +
  labs(title = "Missed Payments by Default Status",
       x = "Default Status", y = "Number of Missed Payments") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Defaulters have a wider distribution and a higher upper quartile of missed payments compared to non-defaulters. This is a strong signal of default risk, indicating users with repeatedly miss payments (e.g., 2 or more) are more likely to default.

```{r repayment-delay, fig.width=4, fig.height=3}
ggplot(df, aes(x = default_flag, y = repayment_delay_days, fill = default_flag)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("No Default" = "#2196F3", "Default" = "#F44336")) +
  labs(title = "Repayment Delay (Days) by Default Status",
       x = "Default Status", y = "Delay in Days") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Defaulters also tend to delay repayments significantly longer than non-defaulters. The higher median repayment delay days among defaulters suggests that late payment behaviour is also a good predictor.

# Debt-to-Income Ratio vs. Default

```{r dti-default, fig.width=4.5, fig.height=3}
ggplot(df, aes(x = default_flag, y = debt_to_income_ratio, fill = default_flag)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("No Default" = "#2196F3", "Default" = "#F44336")) +
  scale_y_continuous(labels = percent) +
  labs(title = "Debt-to-Income Ratio by Default Status",
       x = "Default Status", y = "Debt-to-Income Ratio") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Defaulters have a significantly higher median debt-to-income ratio compared to non-defaulters. Users spending a larger proportion of their income on debt repayments are correlated with higher financial risk.

# Default Rate by Product Category

```{r product-default, fig.width=5, fig.height=3}
df %>%
  group_by(product_category) %>%
  summarise(
    total   = n(),
    default = sum(default_flag == "Default"),
    rate    = default / total
  ) %>%
  ggplot(aes(x = reorder(product_category, -rate), y = rate, fill = product_category)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_y_continuous(labels = percent, limits = c(0, 0.55)) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Default Rate by Product Category",
       x = "Product Category", y = "Default Rate") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Default rates across product categories are relatively similar, ranging from 37.7% (Sports) to 41.0% (Fashion). While Fashion purchases show a slightly higher default rate, the differences are not dramatic.

# Correlation Heatmap of Numerical Variables

```{r correlation-heatmap, fig.width=5.5, fig.height=4}
cor_data <- df %>%
  select(age, monthly_income, credit_score, purchase_amount,
         repayment_delay_days, missed_payments,
         debt_to_income_ratio, risk_score, app_usage_frequency) %>%
  cor()

corrplot(cor_data,
         method      = "color",
         type        = "upper",
         addCoef.col = "black",
         tl.col      = "black",
         tl.srt      = 45,
         number.cex  = 0.6,
         tl.cex      = 0.7,
         col         = colorRampPalette(c("#F44336", "white", "#2196F3"))(200),
         title       = "Correlation Matrix",
         mar         = c(0, 0, 1, 0))
```

**Interpretation:** The correlation matrix reveals several notable relationships.

- **repayment_delay_days** is positively correlated with **missed_payments**, meaning that the longer a user delays a payment, the more likely the transaction will be flagged as a missed payment. 
- **credit_score** is negatively correlated with **debt_to_income_ratio**, suggesting that users with heavier debt burdens are perceived as higher risk. 
- Notably, **monthly_income** and **credit_score** have an extremely high correlation, indicating severe multicollinearity; **monthly_income** is therefore removed from modelling. 
- Furthermore, **risk_score** shows abnormally strong correlations with multiple variables (e.g., r = -0.89 with **credit_score**), which raises potential data leakage concerns and should be removed.
