---
title: "BNPL Credit Risk — Modeling (Logistic Regression and k-NN)"
subtitle: "ISOM3018 Course Project"
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 6.5,
  fig.height = 4
)
```

```{r libraries}
library(tidymodels)
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r load-data}
train_df <- readRDS("train_df.rds")
test_df  <- readRDS("test_df.rds")

nm_tr <- names(train_df)
nm_te <- names(test_df)
if (!setequal(nm_tr, nm_te)) {
  stop("Column mismatch between train and test sets. Recreate both files by running the preprocessing script.")
}
if (!identical(nm_tr, nm_te)) {
  test_df <- test_df %>% select(all_of(nm_tr))
}

event_lvl <- "second"
```

\newpage

# Model 1: Logistic Regression

The logistic regression model estimates the log-odds of default using the linear combination of input features. It can provide an interpretation of how each variable influences risk. To allow comparability across different independent variables, all numeric variables are standardized with a mean of 0 and a standard deviation of 1.

```{r logistic-fit}
rec_logistic <- recipe(default_flag ~ ., data = train_df) %>%
  step_normalize(all_numeric_predictors())

lr_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

wf_logistic <- workflow() %>%
  add_recipe(rec_logistic) %>%
  add_model(lr_spec)

set.seed(42)
fit_logistic <- fit(wf_logistic, data = train_df)
```

```{r logistic-coef-table}
coef_tbl <- tidy(fit_logistic) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate))) %>%
  mutate(
    estimate  = round(estimate, 4),
    std.error = round(std.error, 4),
    p.value   = round(p.value, 4)
  ) %>%
  rename(
    Predictor   = term,
    Coefficient = estimate,
    `Std. Error` = std.error,
    `p-value`   = p.value
  ) %>%
  select(Predictor, Coefficient, `Std. Error`, `p-value`)

kable(
  coef_tbl,
  booktabs = TRUE,
  caption  = "Logistic Regression Coefficients"
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

With 95% confidence level, `repayment_delay_days`, `purchase_amount`, and `missed_payments` have positive coefficients. These variables relate to higher default risk. While `credit_score`, `debt_to_income_ratio`, `employment_type_Unemployed`, and `employment_type_Student` have negative coefficients, indicating lower default risk. Among them, `credit_score` and `repayment_delay_days` have the largest impact on default risk.

```{r logistic-predictions}
test_aug_lr  <- augment(fit_logistic, test_df)
pred_cols_lr <- names(test_aug_lr)[startsWith(names(test_aug_lr), ".pred_")]

test_aug_lr %>%
  mutate(row_id = row_number(), .before = 1) %>%
  select(row_id, default_flag, all_of(pred_cols_lr)) %>%
  write_csv("predictions_logistic_test.csv")
```

\newpage

# Model 2: k-Nearest Neighbors (k-NN)

k-NN classifies each customer based on the majority default outcome among their *k* nearest neighbors. All numeric predictors are standardized before computing distances. The optimal *k* is selected by 5-fold cross-validation on the training data to maximize the ROC AUC. Here, the *k* values from 3 to 101 are evaluated to reduce the risk of ties.

```{r knn-tune}
rec_knn <- recipe(default_flag ~ ., data = train_df) %>%
  step_normalize(all_numeric_predictors())

knn_spec <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("classification")

wf_knn <- workflow() %>%
  add_recipe(rec_knn) %>%
  add_model(knn_spec)

k_grid <- tibble(neighbors = seq(3, 101, by = 2))

set.seed(42)
folds <- vfold_cv(train_df, v = 5, strata = default_flag)

set.seed(42)
tune_res <- tune_grid(
  wf_knn,
  resamples = folds,
  grid      = k_grid,
  metrics   = metric_set(roc_auc, accuracy),
  control   = control_grid(save_pred = FALSE, verbose = FALSE)
)

best_k <- tune_res %>% select_best(metric = "roc_auc")

cv_raw <- tune_res %>%
  collect_metrics() %>%
  dplyr::filter(.metric == "roc_auc")

estimate_col <- if ("mean" %in% names(cv_raw)) "mean" else ".estimate"

cv_k_auc <- cv_raw %>%
  dplyr::mutate(estimate_metric = .data[[estimate_col]]) %>%
  dplyr::group_by(neighbors) %>%
  dplyr::summarise(mean_roc_auc = mean(estimate_metric, na.rm = TRUE), .groups = "drop")
```

```{r knn-bestk-table}
kable(
  tibble(`Selected k` = best_k$neighbors,
         `Selection Criterion` = "Max CV ROC AUC",
         `CV Folds` = 5,
         `k Range Searched` = "3 to 101 (odd values only)"),
  booktabs = TRUE,
  caption  = "Optimal k Selected by Cross-validation"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r knn-cv-plot, fig.cap = "Cross-validated ROC AUC by k (Training Data Only)."}
ggplot(cv_k_auc, aes(x = neighbors, y = mean_roc_auc)) +
  geom_line(linewidth = 0.7, color = "#2E7D32") +
  geom_point(size = 1.5, color = "#2E7D32") +
  geom_vline(xintercept = best_k$neighbors, linetype = "dashed", color = "grey35") +
  annotate("text",
           x     = best_k$neighbors + 2,
           y     = min(cv_k_auc$mean_roc_auc),
           label = paste0("k = ", best_k$neighbors),
           hjust = 0, size = 3.2, color = "grey35") +
  labs(x = "k (Number of Neighbors)", y = "Mean CV ROC AUC") +
  theme_minimal(base_size = 11)
```

We can observe that the ROC AUC peaks at k = `r best_k$neighbors`, so *k* is set to be `r best_k$neighbors`.

```{r knn-final-fit}
wf_knn_final <- finalize_workflow(wf_knn, best_k)
set.seed(42)
fit_knn <- fit(wf_knn_final, data = train_df)
```

```{r knn-predictions}
test_aug_knn  <- augment(fit_knn, test_df)
pred_cols_knn <- names(test_aug_knn)[startsWith(names(test_aug_knn), ".pred_")]

test_aug_knn %>%
  mutate(row_id = row_number(), .before = 1) %>%
  select(row_id, default_flag, all_of(pred_cols_knn)) %>%
  write_csv("predictions_knn_test.csv")
```

