---
title: "BNPL Credit Risk — Modeling (Logistic Regression and k-NN)"
subtitle: "ISOM3018 Course Project"
output:
  pdf_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 6.5,
  fig.height = 4
)
```

```{r libraries}
library(tidymodels)
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
```

# Section 1: Modeling Overview

Two classification models — logistic regression and k-nearest neighbors (k-NN) — are built to predict whether a BNPL customer will default on their repayment (`default_flag`: Default vs. No Default).

Both models are trained on the same preprocessed training set (SMOTE applied to address class imbalance) and evaluated on a held-out test set. Numeric predictors are standardized before fitting. The positive class for all discrimination metrics is **Default**.

\newpage

# Section 2: Logistic Regression

```{r load-data}
train_df <- readRDS("train_df.rds")
test_df  <- readRDS("test_df.rds")

nm_tr <- names(train_df)
nm_te <- names(test_df)
if (!setequal(nm_tr, nm_te)) {
  stop("Column mismatch between train and test sets. Recreate both files by running the preprocessing script.")
}
if (!identical(nm_tr, nm_te)) {
  test_df <- test_df %>% select(all_of(nm_tr))
}

event_lvl <- "second"
```

```{r logistic-spec}
rec_logistic <- recipe(default_flag ~ ., data = train_df) %>%
  step_normalize(all_numeric_predictors())

lr_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

wf_logistic <- workflow() %>%
  add_recipe(rec_logistic) %>%
  add_model(lr_spec)

set.seed(42)
fit_logistic <- fit(wf_logistic, data = train_df)

test_aug_lr <- augment(fit_logistic, test_df)

pred_cols_lr <- names(test_aug_lr)[startsWith(names(test_aug_lr), ".pred_")]
pred_export_lr <- test_aug_lr %>%
  mutate(row_id = row_number(), .before = 1) %>%
  select(row_id, default_flag, all_of(pred_cols_lr))
write_csv(pred_export_lr, "predictions_logistic_test.csv")

m_lr <- bind_rows(
  roc_auc(test_aug_lr, truth = default_flag, .pred_Default, event_level = event_lvl),
  accuracy(test_aug_lr, truth = default_flag, .pred_class),
  yardstick::sens(test_aug_lr, truth = default_flag, estimate = .pred_class, event_level = event_lvl),
  yardstick::spec(test_aug_lr, truth = default_flag, estimate = .pred_class, event_level = event_lvl)
) %>%
  mutate(model = "Logistic Regression", .before = 1)

cm_lr  <- conf_mat(test_aug_lr, truth = default_flag, estimate = .pred_class)
roc_lr <- roc_curve(test_aug_lr, default_flag, .pred_Default, event_level = event_lvl)
```

Logistic regression models the probability of default as a linear function of customer characteristics. It is selected as a baseline for its interpretability — the contribution of each predictor can be directly examined. All numeric predictors are standardized; categorical predictors are dummy-encoded.

```{r logistic-metrics-table}
kable(
  m_lr %>%
    dplyr::rename(Metric = `.metric`, Value = `.estimate`) %>%
    dplyr::select(Metric, Value),
  digits = 4,
  booktabs = TRUE,
  caption = "Test Set Performance — Logistic Regression"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

**Interpretation:** The model achieves a ROC AUC of 0.74, indicating moderate discriminative ability. At the default 0.50 threshold, it correctly identifies around 65% of actual defaulters (sensitivity), while maintaining a similar rate for non-defaulters (specificity). For a BNPL provider, higher sensitivity is generally preferred — missing a defaulter is more costly than a false alarm.

```{r logistic-roc, fig.cap = "ROC Curve — Logistic Regression (Test Set)."}
ggplot(roc_lr, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path(linewidth = 0.8, color = "#1565C0") +
  geom_abline(linetype = "dashed", color = "grey50") +
  coord_equal() +
  labs(x = "False Positive Rate", y = "True Positive Rate (Sensitivity)") +
  theme_minimal(base_size = 11)
```

**Interpretation:** The ROC curve lies clearly above the diagonal (random baseline), confirming the model has meaningful predictive power for identifying BNPL default risk.

```{r logistic-confusion}
cm_tbl_lr <- cm_lr$table

cm_display_lr <- tibble(
  ` ` = c("Actual: Default", "Actual: No Default"),
  `Predicted: Default`    = c(cm_tbl_lr["Default",    "Default"],
                               cm_tbl_lr["No_Default", "Default"]),
  `Predicted: No Default` = c(cm_tbl_lr["Default",    "No_Default"],
                               cm_tbl_lr["No_Default", "No_Default"])
)

kable(
  cm_display_lr,
  booktabs = TRUE,
  caption  = "Confusion Matrix — Logistic Regression (Test Set)"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  column_spec(1, bold = TRUE)
```

**Interpretation:** Of the 808 customers who actually defaulted, 523 were correctly flagged (true positives), while 285 were missed (false negatives). Of the 1,261 non-defaulters, 844 were correctly cleared and 417 were incorrectly flagged. Reducing the false negative count — undetected defaulters — is the primary operational priority for credit risk management.

\newpage

# Section 3: k-Nearest Neighbors (k-NN)

```{r knn-tune}
rec_knn <- recipe(default_flag ~ ., data = train_df) %>%
  step_normalize(all_numeric_predictors())

knn_spec <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("classification")

wf_knn <- workflow() %>%
  add_recipe(rec_knn) %>%
  add_model(knn_spec)

k_grid <- tibble(neighbors = seq(3, 101, by = 2))

set.seed(42)
folds <- vfold_cv(train_df, v = 5, strata = default_flag)

set.seed(42)
tune_res <- tune_grid(
  wf_knn,
  resamples = folds,
  grid      = k_grid,
  metrics   = metric_set(roc_auc, accuracy),
  control   = control_grid(save_pred = FALSE, verbose = FALSE)
)

best_k <- tune_res %>% select_best(metric = "roc_auc")

wf_knn_final <- finalize_workflow(wf_knn, best_k)
set.seed(42)
fit_knn <- fit(wf_knn_final, data = train_df)

test_aug_knn <- augment(fit_knn, test_df)

pred_cols_knn <- names(test_aug_knn)[startsWith(names(test_aug_knn), ".pred_")]
pred_export_knn <- test_aug_knn %>%
  mutate(row_id = row_number(), .before = 1) %>%
  select(row_id, default_flag, all_of(pred_cols_knn))
write_csv(pred_export_knn, "predictions_knn_test.csv")

m_knn <- bind_rows(
  roc_auc(test_aug_knn, truth = default_flag, .pred_Default, event_level = event_lvl),
  accuracy(test_aug_knn, truth = default_flag, .pred_class),
  yardstick::sens(test_aug_knn, truth = default_flag, estimate = .pred_class, event_level = event_lvl),
  yardstick::spec(test_aug_knn, truth = default_flag, estimate = .pred_class, event_level = event_lvl)
) %>%
  mutate(model = "k-NN", .before = 1)

cm_knn  <- conf_mat(test_aug_knn, truth = default_flag, estimate = .pred_class)
roc_knn <- roc_curve(test_aug_knn, default_flag, .pred_Default, event_level = event_lvl)

cv_raw <- tune_res %>%
  collect_metrics() %>%
  dplyr::filter(.metric == "roc_auc")

estimate_col <- if (".estimate" %in% names(cv_raw)) {
  ".estimate"
} else if ("mean" %in% names(cv_raw)) {
  "mean"
} else {
  stop("collect_metrics() missing a usable value column. Available: ", paste(names(cv_raw), collapse = ", "))
}

cv_k_auc <- cv_raw %>%
  dplyr::mutate(estimate_metric = .data[[estimate_col]]) %>%
  dplyr::group_by(neighbors) %>%
  dplyr::summarise(mean_roc_auc = mean(estimate_metric, na.rm = TRUE), .groups = "drop")
```

k-NN classifies each customer based on the majority default outcome among their *k* nearest neighbors in standardized feature space. The optimal *k* is selected by 5-fold cross-validation on the training data, choosing the value that maximizes ROC AUC.

```{r knn-bestk-table}
kable(
  tibble(`Selected k` = best_k$neighbors,
         `Selection Criterion` = "Max CV ROC AUC"),
  booktabs = TRUE,
  caption  = "Optimal k Selected by Cross-validation"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r knn-cv-plot, fig.cap = "Cross-validated ROC AUC by k (Training Data Only)."}
ggplot(cv_k_auc, aes(x = neighbors, y = mean_roc_auc)) +
  geom_line(linewidth = 0.7, color = "#2E7D32") +
  geom_point(size = 2, color = "#2E7D32") +
  geom_vline(xintercept = best_k$neighbors, linetype = "dashed", color = "grey35") +
  labs(x = "k (Number of Neighbors)", y = "Mean CV ROC AUC") +
  theme_minimal(base_size = 11)
```

**Interpretation:** ROC AUC increases steadily as *k* grows, peaking at k = `r best_k$neighbors`. A larger *k* produces smoother, more stable predictions by averaging over more neighbors, which helps on this dataset.

```{r knn-metrics-table}
kable(
  m_knn %>%
    dplyr::rename(Metric = `.metric`, Value = `.estimate`) %>%
    dplyr::select(Metric, Value),
  digits = 4,
  booktabs = TRUE,
  caption  = "Test Set Performance — k-NN"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

**Interpretation:** k-NN achieves a ROC AUC of 0.72, slightly below logistic regression. Its sensitivity (true positive rate for defaulters) is 0.62, meaning it detects fewer defaulters compared to logistic regression. However, it achieves a marginally higher specificity, correctly clearing more non-defaulters.

```{r knn-roc, fig.cap = "ROC Curve — k-NN (Test Set)."}
ggplot(roc_knn, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path(linewidth = 0.8, color = "#2E7D32") +
  geom_abline(linetype = "dashed", color = "grey50") +
  coord_equal() +
  labs(x = "False Positive Rate", y = "True Positive Rate (Sensitivity)") +
  theme_minimal(base_size = 11)
```

```{r knn-confusion}
cm_tbl_knn <- cm_knn$table

cm_display_knn <- tibble(
  ` ` = c("Actual: Default", "Actual: No Default"),
  `Predicted: Default`    = c(cm_tbl_knn["Default",    "Default"],
                               cm_tbl_knn["No_Default", "Default"]),
  `Predicted: No Default` = c(cm_tbl_knn["Default",    "No_Default"],
                               cm_tbl_knn["No_Default", "No_Default"])
)

kable(
  cm_display_knn,
  booktabs = TRUE,
  caption  = "Confusion Matrix — k-NN (Test Set)"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  column_spec(1, bold = TRUE)
```

**Interpretation:** k-NN correctly identifies 502 defaulters (true positives) but misses 414 (false negatives). Compared to logistic regression, it catches fewer actual defaulters — a disadvantage in credit risk contexts where undetected defaults carry higher financial cost.

\newpage

# Section 4: Model Comparison

```{r compare-table}
compare_tbl <- bind_rows(m_lr, m_knn) %>%
  dplyr::rename(metric_name = `.metric`, metric_value = `.estimate`) %>%
  dplyr::select(model, metric_name, metric_value) %>%
  tidyr::pivot_wider(names_from = metric_name, values_from = metric_value) %>%
  dplyr::rename(
    Model       = model,
    `ROC AUC`   = roc_auc,
    Accuracy    = accuracy,
    Sensitivity = sens,
    Specificity = spec
  )

kable(compare_tbl, digits = 4, booktabs = TRUE,
      caption = "Test Set Performance — Logistic Regression vs. k-NN") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 10)
```

```{r compare-roc, fig.cap = "ROC Curves — Logistic Regression vs. k-NN (Test Set)."}
bind_rows(
  roc_lr  %>% mutate(model = "Logistic Regression"),
  roc_knn %>% mutate(model = "k-NN")
) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(linewidth = 0.8) +
  geom_abline(linetype = "dashed", color = "grey50") +
  coord_equal() +
  scale_color_manual(values = c("Logistic Regression" = "#1565C0", "k-NN" = "#2E7D32")) +
  labs(x = "False Positive Rate", y = "True Positive Rate", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Interpretation:** Logistic regression outperforms k-NN on ROC AUC (0.74 vs. 0.72), accuracy, and sensitivity. Both models achieve similar specificity. For BNPL default detection, logistic regression is the stronger performer of the two.

\newpage

# Section 5: Business Implications

- **Logistic regression** is the better-performing model of the two and has the added advantage of interpretability — each predictor's effect on default probability can be quantified and explained to risk managers and regulators.
- **k-NN** performs slightly below logistic regression on most metrics. Its lower sensitivity means more defaults go undetected, which is a significant drawback in a credit risk context.
- At the current 0.50 probability threshold, both models miss roughly a third of actual defaulters. A BNPL provider could lower the threshold to flag more potential defaulters, trading some specificity for higher sensitivity — a business decision depending on the relative cost of missed defaults versus false alarms.
- The predicted default probabilities from both models can support practical risk actions: prioritizing manual review for borderline cases, applying lower credit limits to high-risk customers, and triggering early repayment reminders.

