---
title: "Random Forest and Decision Tree"
output: pdf_document
date: "2026-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```


```{r}
library(randomForest)
library(rpart)
library(rpart.plot)
library(caret)
library(pROC)


```

# 1. Model Development

Two models were developed to predict customer default risk: Decision Tree and Random Forest. These models were selected to compare interpretability and predictive performance.
```{r}
# Load data
train_df <- readRDS("train_df.rds")
test_df  <- readRDS("test_df.rds")

validation_df <- test_df

as_binary_factor <- function(y) {
  if (is.numeric(y) || is.integer(y)) {
    y <- factor(y, levels = sort(unique(y)))
  } else if (!is.factor(y)) {
    y <- factor(y)
  }
  if (nlevels(y) != 2) {
    stop("default_flag must be a binary outcome (2 levels).")
  }
  y
}

eval_classifier <- function(truth, pred_class, pred_prob) {
  truth <- as_binary_factor(truth)
  pred_class <- factor(pred_class, levels = levels(truth))
  positive_class <- levels(truth)[2]

  conf <- confusionMatrix(pred_class, truth, positive = positive_class)
  roc_obj <- roc(truth, pred_prob, levels = levels(truth), quiet = TRUE)

  list(
    conf = conf,
    roc = roc_obj,
    auc = as.numeric(auc(roc_obj)),
    metrics = data.frame(
      Accuracy = as.numeric(conf$overall["Accuracy"]),
      Sensitivity = as.numeric(conf$byClass["Sensitivity"]),
      Specificity = as.numeric(conf$byClass["Specificity"]),
      Recall = as.numeric(conf$byClass["Recall"]),
      Precision = as.numeric(conf$byClass["Precision"]),
      F1 = as.numeric(conf$byClass["F1"]),
      AUC = as.numeric(auc(roc_obj))
    )
  )
}

train_df$default_flag <- as_binary_factor(train_df$default_flag)
validation_df$default_flag <- as_binary_factor(validation_df$default_flag)

validation_y <- validation_df$default_flag
positive_class <- levels(validation_y)[2]

```
# 2. Decision Tree Model

```{r}
set.seed(42)

# 2.1 Default tree (rpart defaults)
dt_default <- rpart(
  default_flag ~ ., 
  data = train_df,
  method = "class"
)

# 2.2 Full tree (grow as large as possible, cp=0)
dt_full <- rpart(
  default_flag ~ .,
  data = train_df,
  method = "class",
  control = rpart.control(cp = 0, minsplit = 2, minbucket = 1, maxdepth = 30, xval = 10)
)

# 2.3 Pruned tree (prune the full tree using the cp with minimum cross-validated error)
cp_table <- dt_full$cptable
best_cp <- cp_table[which.min(cp_table[, "xerror"]), "CP"]
dt_pruned <- prune(dt_full, cp = best_cp)
```



## 2.1 Tree Visualization

```{r, fig.show='hold', fig.width=7, fig.height=4}
rpart.plot(dt_default, main = "Decision Tree (Default)")
rpart.plot(dt_pruned, main = "Decision Tree (Pruned)")


```

## 2.2 Tree Complexity (Leaf Count)
```{r}
count_leaves <- function(fit) sum(fit$frame$var == "<leaf>")
count_splits <- function(fit) sum(fit$frame$var != "<leaf>")
count_nodes <- function(fit) nrow(fit$frame)

leaf_counts <- data.frame(
  Model = c("Decision Tree (Default)", "Decision Tree (Full)", "Decision Tree (Pruned)"),
  Leaves = c(count_leaves(dt_default), count_leaves(dt_full), count_leaves(dt_pruned)),
  Splits = c(count_splits(dt_default), count_splits(dt_full), count_splits(dt_pruned)),
  Nodes = c(count_nodes(dt_default), count_nodes(dt_full), count_nodes(dt_pruned))
)

knitr::kable(leaf_counts, caption = "Decision Tree Complexity (Leaf Count)")
```

```{r}
# Validation-set predictions + evaluation
dt_default_pred_class <- predict(dt_default, validation_df, type = "class")
dt_default_pred_prob  <- predict(dt_default, validation_df, type = "prob")[, positive_class]

dt_full_pred_class <- predict(dt_full, validation_df, type = "class")
dt_full_pred_prob  <- predict(dt_full, validation_df, type = "prob")[, positive_class]

dt_pruned_pred_class <- predict(dt_pruned, validation_df, type = "class")
dt_pruned_pred_prob  <- predict(dt_pruned, validation_df, type = "prob")[, positive_class]

dt_default_eval <- eval_classifier(validation_y, dt_default_pred_class, dt_default_pred_prob)
dt_full_eval <- eval_classifier(validation_y, dt_full_pred_class, dt_full_pred_prob)
dt_pruned_eval <- eval_classifier(validation_y, dt_pruned_pred_class, dt_pruned_pred_prob)

dt_results <- rbind(
  data.frame(Model = "Decision Tree (Default)", dt_default_eval$metrics, check.names = FALSE),
  data.frame(Model = "Decision Tree (Full)", dt_full_eval$metrics, check.names = FALSE),
  data.frame(Model = "Decision Tree (Pruned)", dt_pruned_eval$metrics, check.names = FALSE)
)
```

## 2.3 Validation Predictions (Sample)
```{r}
dt_validation_predictions <- data.frame(
  Actual = validation_y,
  DT_Default_Class = dt_default_pred_class,
  DT_Default_Prob = round(dt_default_pred_prob, 4),
  DT_Full_Class = dt_full_pred_class,
  DT_Full_Prob = round(dt_full_pred_prob, 4),
  DT_Pruned_Class = dt_pruned_pred_class,
  DT_Pruned_Prob = round(dt_pruned_pred_prob, 4)
)

knitr::kable(head(dt_validation_predictions, 10), caption = "Decision Tree Predictions (First 10 Validation Rows)")
```


## 2.4 Performance
Decision tree performance on the validation set:
```{r}
knitr::kable(dt_results, digits = 3, caption = "Decision Tree Performance (Validation Set)")

knitr::kable(dt_default_eval$conf$table, caption = "Confusion Matrix - Decision Tree (Default)")
knitr::kable(dt_full_eval$conf$table, caption = "Confusion Matrix - Decision Tree (Full)")
knitr::kable(dt_pruned_eval$conf$table, caption = "Confusion Matrix - Decision Tree (Pruned)")

plot(dt_default_eval$roc, col = "red", main = "ROC Curves - Decision Trees")
plot(dt_full_eval$roc, col = "darkorange", add = TRUE)
plot(dt_pruned_eval$roc, col = "darkgreen", add = TRUE)
legend(
  "bottomright",
  legend = c("Default", "Full", "Pruned"),
  col = c("red", "darkorange", "darkgreen"),
  lwd = 2,
  bty = "n"
)

plotcp(dt_full)
```


# 3. Random Forest Model


```{r}
set.seed(42)

# 3.1 Random Forest (Default settings)
rf_default <- randomForest(
  default_flag ~ .,
  data = train_df,
  importance = TRUE
)

# 3.2 Random Forest (Tuned mtry using OOB error)
train_y <- as_binary_factor(train_df$default_flag)
train_x <- model.matrix(default_flag ~ ., data = train_df)[, -1, drop = FALSE]

tuned <- tuneRF(
  x = train_x,
  y = train_y,
  ntreeTry = 500,
  stepFactor = 1.5,
  improve = 0.01,
  trace = FALSE,
  plot = FALSE
)
best_mtry <- tuned[which.min(tuned[, "OOBError"]), "mtry"]

rf_tuned <- randomForest(
  default_flag ~ .,
  data = train_df,
  ntree = 500,
  mtry = best_mtry,
  importance = TRUE
)

# 3.3 Random Forest (All features per split; bagging-style)
rf_all_features <- randomForest(
  default_flag ~ .,
  data = train_df,
  ntree = 500,
  mtry = ncol(train_x),
  importance = TRUE
)
```



```{r}
# Validation-set prediction & evaluation
rf_default_pred_class <- predict(rf_default, validation_df)
rf_default_pred_prob  <- predict(rf_default, validation_df, type = "prob")[, positive_class]
rf_default_eval <- eval_classifier(validation_y, rf_default_pred_class, rf_default_pred_prob)

rf_tuned_pred_class <- predict(rf_tuned, validation_df)
rf_tuned_pred_prob  <- predict(rf_tuned, validation_df, type = "prob")[, positive_class]
rf_tuned_eval <- eval_classifier(validation_y, rf_tuned_pred_class, rf_tuned_pred_prob)

rf_all_pred_class <- predict(rf_all_features, validation_df)
rf_all_pred_prob  <- predict(rf_all_features, validation_df, type = "prob")[, positive_class]
rf_all_eval <- eval_classifier(validation_y, rf_all_pred_class, rf_all_pred_prob)

rf_results <- rbind(
  data.frame(Model = "Random Forest (Default)", rf_default_eval$metrics, check.names = FALSE),
  data.frame(Model = "Random Forest (Tuned mtry)", rf_tuned_eval$metrics, check.names = FALSE),
  data.frame(Model = "Random Forest (All features)", rf_all_eval$metrics, check.names = FALSE)
)
```

## 3.1 Performance

Random forest performance on the validation set:
```{r}
knitr::kable(rf_results, digits = 3, caption = "Random Forest Performance (Validation Set)")

knitr::kable(rf_default_eval$conf$table, caption = "Confusion Matrix - Random Forest (Default)")
knitr::kable(rf_tuned_eval$conf$table, caption = "Confusion Matrix - Random Forest (Tuned mtry)")
knitr::kable(rf_all_eval$conf$table, caption = "Confusion Matrix - Random Forest (All features)")

plot(rf_default_eval$roc, col = "blue", main = "ROC Curves - Random Forest")
plot(rf_tuned_eval$roc, col = "purple", add = TRUE)
plot(rf_all_eval$roc, col = "darkcyan", add = TRUE)
legend(
  "bottomright",
  legend = c("Default", "Tuned mtry", "All features"),
  col = c("blue", "purple", "darkcyan"),
  lwd = 2,
  bty = "n"
)
```

## 3.2 Validation Predictions (Sample)
```{r}
rf_validation_predictions <- data.frame(
  Actual = validation_y,
  RF_Default_Class = rf_default_pred_class,
  RF_Default_Prob = round(rf_default_pred_prob, 4),
  RF_Tuned_Class = rf_tuned_pred_class,
  RF_Tuned_Prob = round(rf_tuned_pred_prob, 4),
  RF_AllFeat_Class = rf_all_pred_class,
  RF_AllFeat_Prob = round(rf_all_pred_prob, 4)
)

knitr::kable(head(rf_validation_predictions, 10), caption = "Random Forest Predictions (First 10 Validation Rows)")
```


## 3.3 Feature Importance
The feature importance below is computed from the best random forest variant by validation AUC.
```{r, fig.align='center', fig.width=6, fig.height=4}
best_rf_idx <- which.max(rf_results$AUC)
best_rf_name <- rf_results$Model[best_rf_idx]
best_rf_model <- list(rf_default, rf_tuned, rf_all_features)[[best_rf_idx]]

# ===== Feature Importance =====
importance_df <- as.data.frame(importance(best_rf_model))

# Convert rownames to column
importance_df$Feature <- rownames(importance_df)

# Keep only what we need
importance_clean <- importance_df[, c("Feature", "MeanDecreaseGini")]

# Sort descending
importance_clean <- importance_clean[order(-importance_clean$MeanDecreaseGini), ]

# Top 10
top_features <- head(importance_clean, 10)

# Remove rownames to avoid duplication
rownames(top_features) <- NULL

# ===== Plot =====
library(ggplot2)

ggplot(top_features, aes(x = reorder(Feature, MeanDecreaseGini), 
                         y = MeanDecreaseGini)) +
  geom_col() +
  coord_flip() +
  labs(
    title = paste0("Top 10 Important Features (", best_rf_name, ")"),
    x = "Feature",
    y = "Importance (Mean Decrease Gini)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# ===== Table =====
knitr::kable(
  top_features,
  digits = 2,
  caption = paste0("Top 10 Important Features (", best_rf_name, ")"),
  row.names = FALSE
)



```
## 3.4 Performance Summary (Best Random Forest)
```{r}
best_rf_eval <- list(rf_default_eval, rf_tuned_eval, rf_all_eval)[[best_rf_idx]]

# ===== Performance Table =====
performance <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "AUC"),
  Value = c(
    best_rf_eval$conf$overall['Accuracy'],
    best_rf_eval$conf$byClass['Sensitivity'],
    best_rf_eval$conf$byClass['Specificity'],
    best_rf_eval$auc
  )
)

# Round values for readability
performance$Value <- round(performance$Value, 3)

knitr::kable(
  performance,
  caption = "Random Forest Performance",
  row.names = FALSE
)
```


# 4. Model Comparison
```{r}
model_results <- rbind(
  dt_results,
  rf_results
)
rownames(model_results) <- NULL

knitr::kable(model_results, digits = 3, caption = "Model Comparison (Validation Set)")

```


# 5. Interpretation

This section explains *why* each modeling step is included, and how to interpret the validation results.

## 5.1 How to read the metrics
We evaluate every model on the **validation set** (`validation_df`, currently set to `test_df`) using:

- **AUC**: overall ranking quality across all probability thresholds (higher is better).
- **Sensitivity (Recall)**: among actual defaulters, how many we correctly flag as default (higher reduces missed high-risk customers).
- **Specificity**: among non-defaulters, how many we correctly approve (higher reduces false alarms / unnecessary rejection).
- **Accuracy**: can be misleading if the classes are imbalanced, so we use it only as a supporting metric.

The best model by validation AUC is **`r model_results$Model[which.max(model_results$AUC)]`** (AUC = `r round(max(model_results$AUC), 3)`).

## 5.2 Decision Tree: why default → full → pruned
**(1) Default tree (`dt_default`)**  
We start with default settings as a baseline: it is fast to train, easy to visualize, and gives an interpretable “first-pass” model. It also helps verify that the data and target are set up correctly.

**(2) Full tree (`dt_full`, cp = 0)**  
We intentionally allow the tree to grow very large to observe the *maximum* complexity the training data can support. This is useful because it typically:
- fits training patterns (including noise) very closely, and
- exposes overfitting risk when validation performance does not improve.

In this run, the full tree has **`r leaf_counts$Leaves[leaf_counts$Model=='Decision Tree (Full)']`** leaves (vs **`r leaf_counts$Leaves[leaf_counts$Model=='Decision Tree (Default)']`** for the default tree).

**(3) Pruned tree (`dt_pruned`)**  
Pruning is used to trade a small amount of training fit for better generalization and simpler rules. Here we prune the full tree using the **complexity parameter** chosen by cross-validation:
- selected `cp` = **`r signif(best_cp, 4)`** (minimum cross-validated error in `dt_full$cptable`)
- pruned tree leaves = **`r leaf_counts$Leaves[leaf_counts$Model=='Decision Tree (Pruned)']`**

If pruning improves or maintains validation AUC while using fewer leaves, it suggests the full tree was capturing noise rather than signal.

## 5.3 Random Forest: why default → tune `mtry` → all-features
**(1) Default random forest (`rf_default`)**  
This is the baseline ensemble. Compared with a single tree, random forests reduce variance by averaging many decorrelated trees.

**(2) Tuned `mtry` (`rf_tuned`)**  
`mtry` controls how many features are randomly considered at each split. Tuning it is a standard step because it changes the diversity/strength trade-off of trees in the ensemble.  
We tune `mtry` using **out-of-bag (OOB) error** via `tuneRF()`, then refit with `ntree = 500`:
- tuned `mtry` = **`r best_mtry`**

**(3) All features per split (`rf_all_features`)**  
Setting `mtry = p` (all predictors) is close to “bagging-style” trees. This is a useful comparison point: if this performs worse than tuned/default RF, it indicates that random feature selection is contributing to improved generalization.

## 5.4 Practical takeaways
- If your goal is to **catch as many defaulters as possible**, prioritize **Sensitivity** (and consider adjusting the decision threshold on predicted probabilities).
- If your goal is to **avoid rejecting good customers**, prioritize **Specificity**.
- Use the **pruned tree** when you need simple, explainable rules; use the **best RF** when you need stronger predictive performance.

# 6. Business Implications

The results suggest that customer default risk is influenced by complex interactions between multiple factors. The Random Forest model can be used to support risk-based decision-making, such as adjusting credit limits or implementing targeted risk controls.

By leveraging these insights, businesses can reduce financial losses while maintaining customer growth through more informed lending strategies.
