---
title: "Random Forest and Decision Tree"
output: pdf_document
date: "2026-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```


```{r}
library(randomForest)
library(rpart)
library(rpart.plot)
library(caret)
library(pROC)


```

# 1. Model Development

Two models were developed to predict customer default risk: Decision Tree and Random Forest. These models were selected to compare interpretability and predictive performance.
```{r}
# Load data
train_df <- readRDS("train_df.rds")
test_df  <- readRDS("test_df.rds")

validation_df <- test_df

as_binary_factor <- function(y) {
  if (is.numeric(y) || is.integer(y)) {
    y <- factor(y, levels = sort(unique(y)))
  } else if (!is.factor(y)) {
    y <- factor(y)
  }
  if (nlevels(y) != 2) {
    stop("default_flag must be a binary outcome (2 levels).")
  }
  y
}

eval_classifier <- function(truth, pred_class, pred_prob) {
  truth <- as_binary_factor(truth)
  pred_class <- factor(pred_class, levels = levels(truth))
  positive_class <- levels(truth)[2]

  conf <- confusionMatrix(pred_class, truth, positive = positive_class)
  roc_obj <- roc(truth, pred_prob, levels = levels(truth), quiet = TRUE)

  list(
    conf = conf,
    roc = roc_obj,
    auc = as.numeric(auc(roc_obj)),
    metrics = data.frame(
      Accuracy = as.numeric(conf$overall["Accuracy"]),
      Sensitivity = as.numeric(conf$byClass["Sensitivity"]),
      Specificity = as.numeric(conf$byClass["Specificity"]),
      Recall = as.numeric(conf$byClass["Recall"]),
      Precision = as.numeric(conf$byClass["Precision"]),
      F1 = as.numeric(conf$byClass["F1"]),
      AUC = as.numeric(auc(roc_obj))
    )
  )
}

train_df$default_flag <- as_binary_factor(train_df$default_flag)
validation_df$default_flag <- as_binary_factor(validation_df$default_flag)

validation_y <- validation_df$default_flag
positive_class <- levels(validation_y)[2]

```

# 2. Decision Tree Model
Decision tree models are being built using rpart to predict the target variable (default_flag). To examine model complexity and potential overfitting issues, three versions of the decision tree are constructed: a default tree, a fully grown tree, and a pruned tree.

Default Tree:
The default decision tree is built using standard rpart parameters, acting as a baseline model.

Full Tree:
A fully grown tree is built by setting the complexity parameter (cp = 0) and allowing minimal restrictions on splitting, capturing all possible patterns in the training data.

Pruned Tree:
In order to solve possible overfitting issues, the full tree is pruned using the optimal complexity parameter (cp) selected by the minimum cross-validated error (xerror). 
```{r}
set.seed(42)

# 2.1 Default tree (rpart defaults)
dt_default <- rpart(
  default_flag ~ ., 
  data = train_df,
  method = "class"
)

# 2.2 Full tree (grow as large as possible, cp=0)
dt_full <- rpart(
  default_flag ~ .,
  data = train_df,
  method = "class",
  control = rpart.control(cp = 0, minsplit = 2, minbucket = 1, maxdepth = 30, xval = 10)
)

# 2.3 Pruned tree (prune the full tree using the cp with minimum cross-validated error)
cp_table <- dt_full$cptable
best_cp <- cp_table[which.min(cp_table[, "xerror"]), "CP"]
dt_pruned <- prune(dt_full, cp = best_cp)
```
## 2.1 Tree Complexity (Leaf Count)
The leaf count of the three trees are being examined to check for complexity and overfitting issues.
```{r}
count_leaves <- function(fit) sum(fit$frame$var == "<leaf>")
count_splits <- function(fit) sum(fit$frame$var != "<leaf>")
count_nodes <- function(fit) nrow(fit$frame)

leaf_counts <- data.frame(
  Model = c("Decision Tree (Default)", "Decision Tree (Full)", "Decision Tree (Pruned)"),
  Leaves = c(count_leaves(dt_default), count_leaves(dt_full), count_leaves(dt_pruned)),
  Splits = c(count_splits(dt_default), count_splits(dt_full), count_splits(dt_pruned)),
  Nodes = c(count_nodes(dt_default), count_nodes(dt_full), count_nodes(dt_pruned))
)

knitr::kable(leaf_counts, caption = "Decision Tree Complexity (Leaf Count)")
```


## 2.2 Tree Visualization
According to the tree complexity results above, the full tree is too deep for visualization and is ineffective to show. Therefore, only the default and pruned tree are being visualized. The visualization shows which combination of the factors will be classified as default or not.

```{r, fig.width=8, fig.height=6}
rpart.plot(
  dt_default,
  main = "Decision Tree (Default)",
  fallen.leaves = TRUE,
  compress = FALSE,
  extra = 0,
  varlen = 0,
  faclen = 0,
  digits = 2,
  cex = 0.7,
  tweak = 0.9,
)
```

```{r, fig.width=20, fig.height=18}
rpart.plot(
  dt_pruned,
  main = "Decision Tree (Pruned)",
  fallen.leaves = FALSE,
  uniform = TRUE,
  compress = FALSE,
  extra = 0,
  varlen = 0,
  faclen = 0,
  digits = 2,
  cex = 0.6,
  tweak = 0.9
)
```


```{r}
# Validation-set predictions + evaluation
dt_default_pred_class <- predict(dt_default, validation_df, type = "class")
dt_default_pred_prob  <- predict(dt_default, validation_df, type = "prob")[, positive_class]

dt_full_pred_class <- predict(dt_full, validation_df, type = "class")
dt_full_pred_prob  <- predict(dt_full, validation_df, type = "prob")[, positive_class]

dt_pruned_pred_class <- predict(dt_pruned, validation_df, type = "class")
dt_pruned_pred_prob  <- predict(dt_pruned, validation_df, type = "prob")[, positive_class]

dt_default_eval <- eval_classifier(validation_y, dt_default_pred_class, dt_default_pred_prob)
dt_full_eval <- eval_classifier(validation_y, dt_full_pred_class, dt_full_pred_prob)
dt_pruned_eval <- eval_classifier(validation_y, dt_pruned_pred_class, dt_pruned_pred_prob)

dt_results <- rbind(
  data.frame(Model = "Decision Tree (Default)", dt_default_eval$metrics, check.names = FALSE),
  data.frame(Model = "Decision Tree (Full)", dt_full_eval$metrics, check.names = FALSE),
  data.frame(Model = "Decision Tree (Pruned)", dt_pruned_eval$metrics, check.names = FALSE)
)
```


## 2.3 Performance
The pruned decision tree provides the best overall performance by having the highest AUC. The default tree shows higher accuracy but performs poorly in identifying default cases due to low sensitivity, while the full tree improves sensitivity but has overfitting problems and weaker overall performance. 
```{r}
knitr::kable(dt_results, digits = 3, caption = "Decision Tree Performance (Validation Set)")

knitr::kable(dt_default_eval$conf$table, caption = "Confusion Matrix - Decision Tree (Default)")
knitr::kable(dt_full_eval$conf$table, caption = "Confusion Matrix - Decision Tree (Full)")
knitr::kable(dt_pruned_eval$conf$table, caption = "Confusion Matrix - Decision Tree (Pruned)")

plot(dt_default_eval$roc, col = "red", main = "ROC Curves - Decision Trees")
plot(dt_full_eval$roc, col = "darkorange", add = TRUE)
plot(dt_pruned_eval$roc, col = "darkgreen", add = TRUE)
legend(
  "bottomright",
  legend = c("Default", "Full", "Pruned"),
  col = c("red", "darkorange", "darkgreen"),
  lwd = 2,
  bty = "n"
)

plotcp(dt_full)
```


# 3. Random Forest Model
Random Forest Models are built to improve predictive performance and reduce overfitting compared to decision trees. Three versions are built: a default model, a tuned model with optimized mtry based on out-of-bag (OOB) error.

```{r}
set.seed(42)

# 3.1 Random Forest (Default settings)
rf_default <- randomForest(
  default_flag ~ .,
  data = train_df,
  importance = TRUE
)

# 3.2 Random Forest (Tuned mtry using OOB error)
train_y <- as_binary_factor(train_df$default_flag)
train_x <- model.matrix(default_flag ~ ., data = train_df)[, -1, drop = FALSE]

tuned <- NULL
suppressWarnings(invisible(capture.output({
  tuned <- tuneRF(
    x = train_x,
    y = train_y,
    ntreeTry = 500,
    stepFactor = 1.5,
    improve = 0.01,
    trace = FALSE,
    plot = FALSE
  )
})))
best_mtry <- tuned[which.min(tuned[, "OOBError"]), "mtry"]

rf_tuned <- randomForest(
  default_flag ~ .,
  data = train_df,
  ntree = 500,
  mtry = best_mtry,
  importance = TRUE
)

# 3.3 Random Forest (All features per split; bagging-style)
rf_all_features <- randomForest(
  default_flag ~ .,
  data = train_df,
  ntree = 500,
  mtry = ncol(train_x),
  importance = TRUE
)
```



```{r}
# Validation-set prediction & evaluation
rf_default_pred_class <- predict(rf_default, validation_df)
rf_default_pred_prob  <- predict(rf_default, validation_df, type = "prob")[, positive_class]
rf_default_eval <- eval_classifier(validation_y, rf_default_pred_class, rf_default_pred_prob)

rf_tuned_pred_class <- predict(rf_tuned, validation_df)
rf_tuned_pred_prob  <- predict(rf_tuned, validation_df, type = "prob")[, positive_class]
rf_tuned_eval <- eval_classifier(validation_y, rf_tuned_pred_class, rf_tuned_pred_prob)

rf_all_pred_class <- predict(rf_all_features, validation_df)
rf_all_pred_prob  <- predict(rf_all_features, validation_df, type = "prob")[, positive_class]
rf_all_eval <- eval_classifier(validation_y, rf_all_pred_class, rf_all_pred_prob)

rf_results <- rbind(
  data.frame(Model = "Random Forest (Default)", rf_default_eval$metrics, check.names = FALSE),
  data.frame(Model = "Random Forest (Tuned mtry)", rf_tuned_eval$metrics, check.names = FALSE),
  data.frame(Model = "Random Forest (All features)", rf_all_eval$metrics, check.names = FALSE)
)
```

## 3.1 Performance
The three random forest models show very similar performance, but the all-features model achieves the highest AUC (0.765), making it the best-performing model overall in terms of classification ability.
```{r}
knitr::kable(rf_results, digits = 3, caption = "Random Forest Performance (Validation Set)")

knitr::kable(rf_default_eval$conf$table, caption = "Confusion Matrix - Random Forest (Default)")
knitr::kable(rf_tuned_eval$conf$table, caption = "Confusion Matrix - Random Forest (Tuned mtry)")
knitr::kable(rf_all_eval$conf$table, caption = "Confusion Matrix - Random Forest (All features)")

plot(rf_default_eval$roc, col = "blue", main = "ROC Curves - Random Forest")
plot(rf_tuned_eval$roc, col = "purple", add = TRUE)
plot(rf_all_eval$roc, col = "darkcyan", add = TRUE)
legend(
  "bottomright",
  legend = c("Default", "Tuned mtry", "All features"),
  col = c("blue", "purple", "darkcyan"),
  lwd = 2,
  bty = "n"
)
```




## 3.2 Feature Importance
The feature importance below is computed from the best random forest chose above. From the results, it shows that the credit_score is the most important feature that affects the result.
```{r, fig.align='center', fig.width=6, fig.height=4}
best_rf_idx <- which.max(rf_results$AUC)
best_rf_name <- rf_results$Model[best_rf_idx]
best_rf_model <- list(rf_default, rf_tuned, rf_all_features)[[best_rf_idx]]

# ===== Feature Importance =====
importance_df <- as.data.frame(importance(best_rf_model))

# Convert rownames to column
importance_df$Feature <- rownames(importance_df)

# Keep only what we need
importance_clean <- importance_df[, c("Feature", "MeanDecreaseGini")]

# Sort descending
importance_clean <- importance_clean[order(-importance_clean$MeanDecreaseGini), ]

# Top 10
top_features <- head(importance_clean, 10)

# Remove rownames to avoid duplication
rownames(top_features) <- NULL

# ===== Plot =====
library(ggplot2)

ggplot(top_features, aes(x = reorder(Feature, MeanDecreaseGini), 
                         y = MeanDecreaseGini)) +
  geom_col() +
  coord_flip() +
  labs(
    title = paste0("Top 10 Important Features"),
    x = "Feature",
    y = "Importance (Mean Decrease Gini)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# ===== Table =====
knitr::kable(
  top_features,
  digits = 2,
  caption = paste0("Top 10 Important Features (", best_rf_name, ")"),
  row.names = FALSE
)



```
## 3.3 Performance Summary (Best Random Forest)
```{r}
best_rf_eval <- list(rf_default_eval, rf_tuned_eval, rf_all_eval)[[best_rf_idx]]

# ===== Performance Table =====
performance <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "AUC"),
  Value = c(
    best_rf_eval$conf$overall['Accuracy'],
    best_rf_eval$conf$byClass['Sensitivity'],
    best_rf_eval$conf$byClass['Specificity'],
    best_rf_eval$auc
  )
)

# Round values for readability
performance$Value <- round(performance$Value, 3)

knitr::kable(
  performance,
  caption = "Random Forest Performance",
  row.names = FALSE
)
```



